---
title: "Price and Volatility Forecasting in Financial Markets"
author: "39603"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
---


```{r setup, include=FALSE}
# Load necessary libraries
library(quantmod)
library(forecast)
library(rugarch)
library(PerformanceAnalytics)
library(ggplot2)
library(knitr)
library(dplyr)

# Set options for better output readability
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction
Forecasting price returns and volatility is essential for risk management, portfolio optimization, and trading strategies. This project focuses on the technology sector, analyzing three prominent stocks: Apple (AAPL), Microsoft (MSFT), and Nvidia (NVDA). It investigates the performance of time series models (ARIMA and GARCH family models) during normal market periods and stress periods (e.g., COVID-19).


# 2. Data Collection and Description
## Data Collection
```{r}

# Define stock symbols and sample period
symbols <- c("AAPL", "MSFT", "NVDA")
start_date <- "2015-01-01"
end_date <- "2024-12-31"

# Download stock data
getSymbols(symbols, src = "yahoo", from = start_date, to = end_date)

# Extract Adjusted Closing Prices
prices <- do.call(merge, lapply(symbols, function(sym) Ad(get(sym))))

# Rename columns
colnames(prices) <- symbols


# Save the prices to a CSV file
#write.csv(as.data.frame(prices), file = "stock_prices.csv", row.names = TRUE)

head(prices,5)

```
## Data Visualization and Returns Calculation
1. **Logarithmic Returns:**
   \[
   R_t = \ln\left(\frac{P_t}{P_{t-1}}\right)
   \]
   where \(R_t\) is the log return, \(P_t\) is the price at time \(t\), and \(P_{t-1}\) is the price at time \(t-1\).


```{r}
# Plot the price trends
autoplot.zoo(prices, facets = NULL) +
  ggtitle("Daily Stock Prices") +
  ylab("Price (USD)") +
  xlab("Date") +
  theme_minimal()
```



```{r}
# Calculate daily log returns
returns <- na.omit(Return.calculate(prices, method = "log"))

# Rename columns for clarity
colnames(returns) <- symbols

# Summary statistics
stats <- data.frame(
  Mean = colMeans(returns),
  SD = apply(returns, 2, sd),
  Skewness = apply(returns, 2, skewness),
  Kurtosis = apply(returns, 2, kurtosis)
)
kable(stats, caption = "Summary Statistics for Daily Returns")
```


```{r}
# Plot returns distribution
autoplot.zoo(returns, facets = NULL) +
  ggtitle("Daily Log Returns") +
  ylab("Returns") +
  xlab("Date") +
  theme_minimal()

```





# 3. Empirical Analysis
## ARIMA for Return Forecasting

2. **ARIMA Model Representation:**
   \[
   \Phi(B)(1 - B)^d X_t = \Theta(B) \epsilon_t
   \]
   where \(B\) is the backshift operator, \(d\) is the differencing order, \(\Phi(B)\) and \(\Theta(B)\) are the polynomials for AR and MA components, and \(\epsilon_t\) is white noise.
   

```{r}
# Fit ARIMA model for Apple returns
apple_returns <- returns$AAPL

arima_model <- auto.arima(apple_returns)
summary(arima_model)

# Forecast returns for the next 30 days
forecasted_returns <- forecast(arima_model, h = 30)

# Plot ARIMA forecast
autoplot(forecasted_returns) +
  ggtitle("ARIMA Forecast for Apple Returns") +
  ylab("Returns") +
  xlab("Date") +
  theme_minimal()

```



## GARCH Family Models for Volatility Forecasting

3. **GARCH(1,1) Model for Volatility:**
   \[
   \sigma_t^2 = \omega + \alpha \epsilon_{t-1}^2 + \beta \sigma_{t-1}^2
   \]
   where \(\sigma_t^2\) is the conditional variance, \(\epsilon_{t-1}^2\) is the squared error term, and \(\omega, \alpha, \beta\) are model parameters.


```{r}
# Define GARCH(1,1) specification
garch_spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0)),
  distribution.model = "norm"
)

# Fit GARCH(1,1) model for Apple returns
garch_fit <- ugarchfit(spec = garch_spec, data = apple_returns)
summary(garch_fit)

garch_fit

# Extract volatility
volatility <- sigma(garch_fit)


# Plot GARCH volatility
plot(volatility, type = "l", col = "blue", main = "GARCH(1,1) Volatility (Apple)", ylab = "Volatility")



```
 


## Compare GARCH, EGARCH, and TGARCH Models

```{r}
# Fit EGARCH model
egarch_spec <- ugarchspec(
  variance.model = list(model = "eGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0)),
  distribution.model = "norm"
)
egarch_fit <- ugarchfit(spec = egarch_spec, data = apple_returns)

# Fit TGARCH model
tgarch_spec <- ugarchspec(
  variance.model = list(model = "fGARCH", garchOrder = c(1, 1), submodel = "TGARCH"),
  mean.model = list(armaOrder = c(0, 0)),
  distribution.model = "norm"
)
tgarch_fit <- ugarchfit(spec = tgarch_spec, data = apple_returns)

# Compare AIC values
model_comparison <- data.frame(
  Model = c("GARCH(1,1)", "EGARCH", "TGARCH"),
  AIC = c(infocriteria(garch_fit)[1], infocriteria(egarch_fit)[1], infocriteria(tgarch_fit)[1])
)
kable(model_comparison, caption = "AIC Comparison of GARCH Family Models")

```
```{r}
# Assuming apple_returns is the data and the models are already fitted

# Set up the plotting area for 3 subplots
par(mfrow = c(1, 3)) 

# Plot GARCH(1,1) model volatility estimate
plot(sigma(garch_fit), type = "l", main = "GARCH(1,1)", col = "blue", xlab = "Time", ylab = "Volatility")

# Plot EGARCH model volatility estimate
plot(sigma(egarch_fit), type = "l", main = "EGARCH", col = "red", xlab = "Time", ylab = "Volatility")

# Plot TGARCH model volatility estimate
plot(sigma(tgarch_fit), type = "l", main = "TGARCH", col = "green", xlab = "Time", ylab = "Volatility")

# Reset the plotting area to default
par(mfrow = c(1, 1))


```


## Rolling Window Analysis
4. **Rolling Window Volatility:**
   \[
   \sigma_t = \sqrt{\frac{1}{n} \sum_{i=t-n+1}^{t} (R_i - \bar{R})^2}
   \]
   where \(n\) is the window size, \(R_i\) are returns, and \(\bar{R}\) is the mean return within the window.


```{r}
# Calculate 30-day rolling volatility
rolling_volatility <- rollapply(returns, width = 30, FUN = sd, by.column = TRUE, fill = NA)

# Plot rolling volatility
autoplot.zoo(rolling_volatility, facets = NULL) +
  ggtitle("30-Day Rolling Volatility") +
  ylab("Volatility") +
  xlab("Date") +
  theme_minimal()


```

```{r}
# Example for AAPL, MSFT, NVDA
stocks <- c("AAPL", "MSFT", "NVDA")  # Replace with actual symbols or data
volatility_stats <- data.frame(Stock = character(0), 
                               Average_Volatility = numeric(0), 
                               Max_Volatility = numeric(0),
                               Min_Volatility = numeric(0))

for(stock in stocks) {
  # Extract the rolling volatility for the stock (replace `returns` with actual stock returns)
  stock_volatility <- rolling_volatility[, stock]  # Adjust if necessary
  
  # Average volatility (2016-2024)
  avg_volatility <- mean(stock_volatility, na.rm = TRUE)
  
  # Max volatility (2020 peak)
  # You would need to filter around 2020; adjust the date range
  max_volatility <- max(stock_volatility['2020'], na.rm = TRUE)
  
  # Min volatility (2022-2024)
  min_volatility <- min(stock_volatility['2022/2024'], na.rm = TRUE)
  
  # Add the result to the dataframe
  volatility_stats <- rbind(volatility_stats, 
                            data.frame(Stock = stock, 
                                       Average_Volatility = avg_volatility, 
                                       Max_Volatility = max_volatility,
                                       Min_Volatility = min_volatility))
}

# Display the statistics
print(volatility_stats)

```


# Backtesting VaR Violations

5. **Value-at-Risk (VaR):**
   \[
   \text{VaR}_{\alpha} = -\text{Quantile}_{\alpha}(R_t)
   \]
   where \(\text{VaR}_{\alpha}\) is the VaR at confidence level \(\alpha\), and \(\text{Quantile}_{\alpha}(R_t)\) represents the \(\alpha\)-quantile of returns.

```{r}
# Calculate Historical VaR (5% confidence level)
VaR <- apply(returns, 2, function(x) quantile(x, probs = 0.05))

# Count violations for Apple stock
apple_violations <- returns$AAPL < VaR["AAPL"]
violation_count <- sum(apple_violations)

# Plot VaR violations
plot(apple_violations, type = "h", col = "red", main = "VaR Violations (Apple)", ylab = "Violations", xlab = "Date")

VaR

```



# 4. Results
ARIMA forecast accuracy is evaluated using MAE and RMSE.
GARCH family models are compared using AIC values.
Rolling volatility highlights clustering during market stress periods.
Backtesting demonstrates how well models capture extreme market events.

# 5. Analysis
Implications: Improved risk management for portfolio strategies.
Broader Context: Insights into regulatory frameworks (e.g., Basel requirements for market risk).
# 6. Conclusion
This study shows that GARCH models perform well for volatility forecasting, particularly in periods of market stress. ARIMA models provide reasonable return forecasts but are less robust during extreme events.


# Bibliography
Danielsson, J. (2011). Financial Risk Forecasting. Wiley.
Bloomberg Terminal Data Access.
R Documentation for rugarch and forecast.

